### 2016-03-28 chapters/background.tex
better integration of the satisfiability section

In the scope of supporting the configuration of a given product line some specific tools and techniques are used. This section gives an overview of what is used in this work to archive the simplification.

\subsection{Feature Models}
Feature models are multi-purpose structure for product lines. 
\begin{figure}
	\includegraphics{img/img-fm.pdf}
	\caption{A simple example of a feature model}
	\label{img-fm}
\end{figure}
One of their benefits is the visualization as a feature diagram, providing an overview of the containing features, and their hierarchy and dependencies. In addition, it classifies the features and their dependencies by Ordinality (Optional/mandatory), logical operator (Disjunction/exclusive disjunction) and whether it is abstract or concrete. Furthermore, it contains all information needed to provide a configuration of Features and its validation.

Although feature models can be represented as a boolean formula in CNF, it is typically represented in form of a feature diagram. The feature diagram is structured as a tree. In that manner feature models map the hierarchy of features. The possible relationships of features with a common parent feature are \textit{or}, \textit{alternative} and \textit{and}. In the case of an \textit{or}-group, at least one of the child features has to be selected if the parent feature is selected. Within an \textit{alternative}-group exactly one of the child feature has to be selected if the parent feature is selected. When grouped with an \textit{and}-relation, any number of child features can be selected, if the parent feature is selected. In addition, child features in an \textit{and}-group can be marked as either mandatory or optional.  As features' relations may be of higher complexity than just parent-child relations additional constraints can be noted within a feature model. Constraints can contain any propositional expression.

Figure~\ref{img-fm} shows an example of a simple feature model. The node \textit{GraphLibrary} is the root feature. Its descendants are the node \textit{Edges}, that has to be selected due to the mandatory marker, and \textit{Algorithms} that is marked optional. The implemented edge types are \textit{Directed} and \textit{Undirected} from which exactly one has to be chosen as they are mutually exclusive. From the algorithms at least one has to be chosen whenever \textit{Algorithms} is selected. The constraint beneath Figure~\ref{img-fm} implies that if the feature \textit{Cycle} is selected, the edge type needs to be directed.

\subsection{Product configuration}
The variability of a product line is represented by its feature model, which itself is a set of features with specific interrelations. The process of configuration of a product line describes the steps to derive a product from the product line. To archive this, a user has to select a subset of all the possible features within the product line to meet his requirements. However, not all subsets of features result in a valid product. The interrelations of the features restrict the possible combinations of features resulting in a \textit{valid} configuration. If only one of the requirements from the interrelations between the features is not met, no product can be created and the configuration as such is considered \textit{invalid}.

During configuration a user selects or deselects features to his needs. This process requires a lot of domain knowledge on the one hand and detailed information about the implementation of each single feature on the other. With growing numbers of features and thus growing numbers of possible products configuration requires more effort. However, the even greater problem arises from the possible interactions and interrelations of features, which one has to keep track of during configuration. The automatic selection or deselection of a features doe to propagation of a made selection always results in a valid configuration. Nevertheless, the propagated selections can be confusing for the user.

@@

\subsection{Satisfiability}
%TODO: rausschmeißen?
A feature model can be expressed as a propositional statement. The mapping between the individual groups (sub-trees) and the corresponding propositional expression is shown in the following list:

{\color{red}TODO: Rules for
\begin{itemize}
	\item Or
	\item Alternative
	\item And
\end{itemize}
}
These particular groups are joined via a logical \textit{AND}. For the example feature model shown in Figure~\ref{img-fm} the corresponding propositional expression is as follows:
\begin{equation}
\begin{split}
	GraphLibrary\ \wedge\ Edges\\
	\wedge\ ((Directed\ \wedge \neg Undirected)\\
	\vee (\neg Directed\ \wedge\ Undirected))
\end{split}
\end{equation}\\

As the feature model not only contains the feature tree, but also additional constraints, these constraints have to be considered in the boolean expression representing the complete feature model. They can be linked to one another via logical \textit{AND}s. The connection to the previously created propositional expression of the feature tree is also made through a logical \textit{AND}. Thus, the complete resulting propositional expression for the feature model shown in Figure~\ref{img-fm} is:
\begin{equation}
\begin{split}
	GraphLibrary\ \wedge\ Edges\\
	\wedge\ ((Directed\ \wedge \neg Undirected)\\
	\vee (\neg Directed\ \wedge\ Undirected))\\
	\wedge\ (Algorithms\ \Rightarrow\ (Number\ \vee\ Cycle))\\
	\wedge\ (Cycle\ \Rightarrow\ Directed)
\end{split}
\end{equation}\\
This formalism allows a configuration to be checked for validity. To do so each selected feature is appended with a logical \textit{AND} and each specifically unselected feature is also appended with a logical \textit{AND} but gets negated. The resulting expression is then evaluated by a SAT-solver to check for satisfiability. Even during configuration this process can be applied to check for invalid partial configurations after each decision.

>>>>

In the scope of supporting the configuration of a given product line some specific tools and techniques are used. This section gives an overview of what is used in this work to archive a simplification of the workflow from a "product line"-like source code base to a final product.

\subsection{Feature Models} \label{ch:fm}
Feature models are multi-purpose structure for product lines. 
\begin{figure}
	\includegraphics{img/img-fm.pdf}
	\caption{A simple example of a feature model}
	\label{img-fm}
\end{figure}
One of their benefits is the visualization as a feature diagram, providing an overview of the containing features, and their hierarchy and dependencies. In addition, it classifies the features and their dependencies by Ordinality (Optional/mandatory), logical operator (Disjunction/exclusive disjunction) and whether it is abstract or concrete. Furthermore, it contains all information needed to provide a configuration of Features and its validation.

Although feature models can be represented as a propositional expression in CNF, it is typically represented in form of a feature diagram. The feature diagram is structured as a tree. In that manner feature models map the hierarchy of features. The possible relationships of features with a common parent feature are \textit{or}, \textit{alternative} and \textit{and}. In the case of an \textit{or}-group, at least one of the child features has to be selected if the parent feature is selected. Within an \textit{alternative}-group exactly one of the child feature has to be selected if the parent feature is selected. When grouped with an \textit{and}-relation, any number of child features can be selected, if the parent feature is selected. In addition, child features in an \textit{and}-group can be marked as either mandatory or optional.  As features' relations may be of higher complexity than just parent-child relations additional constraints can be noted within a feature model. Constraints can contain any propositional expression.

Figure~\ref{img-fm} shows an example of a simple feature model. The node \textit{GraphLibrary} is the root feature. It's childs are the node \textit{Edges}, that has to be selected due to the mandatory marker, and \textit{Algorithms} that is marked optional. The implemented edge types are \textit{Directed} and \textit{Undirected} from which exactly one has to be chosen as they are mutually exclusive. From the algorithms at least one has to be chosen whenever \textit{Algorithms} is selected. The constraint beneath Figure~\ref{img-fm} implies that if the feature \textit{Cycle} is selected, the edge type needs to be directed.

\subsection{Product configuration}
The variability of a product line is represented by it's feature model, which itself is a set of features with specific interrelations. The process of configuration of a product line describes the steps to derive a product from the product line. To archive this, a user has to select a subset of all the possible features within the product line to meet his requirements. However, not all subsets of features result in a valid product. The interrelations of the features restrict the possible combinations of features resulting in a \textit{valid} configuration. If only one of the requirements from the interrelations between the features is validated, no product can be created and the configuration as such is considered \textit{invalid}. For automatic detection of validity, check the next section.

@@

During configuration a user selects or deselects features to his needs. This process requires a lot of domain knowledge on the one hand and detailed information about the implementation of each single feature on the other. With growing numbers of features and thus growing numbers of possible products, configuration requires more effort. However, the even greater problem arises from the possible interactions and interrelations of features, which one has to keep track of during configuration.

For all the interrelations of features which are contained within the feature model, an automated observation of the resulting constraints supports configuration. It does so by checking for satisfiability after each change in the configuration.

\subsection{Satisfiability} \label{ch:sat}
During creation of the feature model as well as during the configuration of a product, validity must always be taken care of. Validity in this context is described as the satisfiability of the corresponding propositional expression.

The formalism of propositional expressions (see Section~\ref{ch:fm} allows feature models and configurations to be checked for validity. The propositional expression of a (partial) configuration is created on the basis of the propositional expression of the underlying feature model. Each selected feature is appended with a logical \textit{AND} and each specifically unselected feature is also appended with a logical \textit{AND} but gets negated. The resulting expression is then evaluated by a SAT-solver to check for satisfiability \cite{sat-solve}.

Even during configuration this process can be applied to check for invalid partial configurations after each decision. If a partial configuration reduces the valid choices in a single group to a single one, it can be made automatically. This is called \textit{propagation}. The automatic selection or deselection of a features due to propagation of a made selection always results in a valid configuration. Nevertheless the propagated selections can be confusing for the user.

### 2016-03-28 chapters/concept.tex
Scenario part 1/2 and small grammar fixes
better integration of the satisfiability section

%unstructured product line $\rightarrow$ generated feature model $\rightarrow$ %optimizing (via SAT-solver?) $\rightarrow$ domain knowledge of an expert %$\Rightarrow$ Questionnaire $\rightarrow$ Feature model + Questionnaire %$\Rightarrow$ Product (variant)\\

@@

The two major steps of extraction of the feature model and the creation and usage of the questionnaire are described in detail in the following sections.

\subsection{Extraction of a Feature Model}
Out of the given structure we extract the hierarchies of the features as well as additional dependencies. These are the input for the automated generation of a feature model. This allows for overview over the product line and it's variability. During our efforts to implement an automated generation of a feature model we encounter some problems for which we have to find solutions/workarounds.

@@

The extracted feature model is also used in the following step of configuration. There, it visualizes the interrelations of the features and supports the understanding of the configuration steps.

@@

This highly depends on the implementation of the questionnaire during development. Our work therefore introduces a set of tools to easily integrate such a questionnaire. The following paragraphs will give an overview over the the possible definitions of a questionnaire.

@@

%TODO: Am ende kommt ein konfiguriertes Produkt raus.

>>>>

@@

The two major steps of extraction of the feature model and the creation and usage of the questionnaire are described in detail in the following sections. As a final result of these steps stands a product targeting the users needs. Supporting the process towards that result is what we aimed at with this work.

\subsection{Extraction of a Feature Model}
Out of the given structure of the source code base we extract the hierarchies of the features as well as additional dependencies. These are the input for the automated generation of a feature model. This allows for overview over the product line and it's variability. During our efforts to implement an automated generation of a feature model we encounter some problems for which we have to find solutions/workarounds.

@@

In each step of the creation of the feature model, e.g. after adding a feature or constraint to the feature model, it is checked by a SAT-solver to avoid creating an invalid feature model which has no valid configurations (see section~\ref{ch:sat}). The extracted feature model is also used in the following step of configuration. There, it visualizes the interrelations of the features and supports the understanding of the configuration steps.

@@

This highly depends on the implementation of the questionnaire during development. Our work therefore introduces a set of tools to easily integrate such a questionnaire. The following paragraphs will give an overview over the possible definitions of a questionnaire.

@@

This allows for experts to design a questionnaire in such a way, that it can guide a user without specific domain knowledge throughout the whole process of configuration. Through just answering questions concerning his specific use-case the user takes all necessary decisions. After he completes the questionnaire, our tool maps his answers to a complete configuration, which is then applied to the existing code artifacts. Finally, a working product is the result.

### 2016-03-28 chapters/conclusion.tex
slight restructuring
TODOs added
better integration of the satisfiability section

\subsection{Conclusion}
This is where the work is concluded. In this section there will be a description of the way we did things and the experiences we made during it. An emphasis will be on the insights and the findings from the scenarios will get outlined.

\subsection{Future Work}
Here will be a summary of the new questions that were raised in this work. Also there will be topics for further research. Particularly the problems we encountered and couldn't solve with our concept and why will be pointed out and first approaches will be suggested.

>>>>

This is where the work is concluded. In this section there will be a description of the way we did things and the experiences we made during it. An emphasis will be on the insights and the findings from the scenarios will get outlined.

Here will be a summary of the new questions that were raised in this work. Also there will be topics for further research. Particularly the problems we encountered and couldn't solve with our concept and why will be pointed out and first approaches will be suggested.

%TODO: research questions
In this work we aim at finding a way to support the process of configuring a software product line. To do so, we firstly introduce feature models as formalism

%TODO: wann ist es sinnvoll, wann nicht?

%TODO: ist das auf andere Projekte übertragbar? Was sind Voraussetzungen?

%TODO: Threats to Validity
